Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-07-24 Writing/manuscript.Rmd
updates

---
title: Aquifer depletion exacerbates agricultural drought losses in the US High Plains
author:
  - Taro Mieno^[Department of Agricultural Economics, University of Nebraska Lincoln, tmieno2@unl.edu]
  - Timothy Foster^[Department of WWW, University of Manchester, timothy.foster@manchester.ac.uk]
  - Shunkei Kakimoto^[Department of Applied Economics, University of Minnesota, kakim002@umn.edu]
  - Nicholas Brozovic^[Department of Agricultural Economics, University of Nebraska Lincoln, nbrozovic@nebraska.edu]
abstract: |
  Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in drought years. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
acknowledgments: |
  Funding for the research in this manuscript was provided by the United States Department of Agriculture under contract numbers OCE 58-0111-20-007 and OCE 58-0111-21-007.
bibliography: DRA.bib
csl: iopart.csl
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    toc: false
    includes:
      in_header: "preamble.tex"
---

@@

```{r}
#| echo: false

@@

```{r run-analys, cache = TRUE, include = FALSE}
# knitr::purl(
#   here::here("GitControlled/Codes/1_1_main_analysis.Rmd"),
#   here::here("GitControlled/Codes/1_1_main_analysis.R"),
# )
# source(here::here("GitControlled/Codes/1_1_main_analysis.R"))

# knitr::purl(
#   here::here("GitControlled/Codes/1_2_sub_analysis.Rmd"),
#   here::here("GitControlled/Codes/1_2_sub_analysis.R"),
# )
# source(here::here("GitControlled/Codes/1_2_sub_analysis.R"))

# knitr::purl(
#   here::here("GitControlled/Codes/2_figures_tables.Rmd"),
#   here::here("GitControlled/Codes/2_figures_tables.R"),
# )
# source(here::here("GitControlled/Codes/2_figures_tables.R"))

# knitr::purl(
#   here::here("GitControlled/Codes/3_appendix.Rmd"),
#   here::here("GitControlled/Codes/3_appendix.R"),
# )

# source(here::here("GitControlled/Codes/3_appendix.R"))
```

@@

Figure \@ref(fig:yield-response) shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (1st row) and soybean (2nd row). In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases. This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors [@zhu2022untangling; @li2020quantifying].

Differentiating irrigated yields by the level of aquifer thickness alters these responses in two ways. First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \@ref(fig:yield-response)). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers ability to fully meet crop water requirements during periods of more severe or extended precipitation deficits [@rad2020effects, @foster2014modeling; @hrozencik2017heterogeneous]. Second, at negative water deficits, counties with the second highest and highest levels of aquifer thickness (i.e., 2nd and 3rd quantile in Figure \@ref(fig:yield-response)) experience lower yields than counties with the lowest level of aquifer thickness (1st quantile), resulting in productivity close to rainfed yields. This counter-intuitive result may be due to the increased possibility of over-irrigation in regions with high aquifer thickness and well capacities [@foster2019assessing; @gibson2017case; @gibson2019benchmarking]. However, it is important to highlight that impacts of aquifer thickness on yields for negative deficits is much less pronounced than that for extreme droughts (i.e. high deficits), in particular for corn.

```{r yield-response, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains", out.height = "5in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_yield_intensive.pdf"))
```

@@

While aquifer thickness has a statistically significant impact on per-area irrigated crop yields in moderate to extreme drought conditions, the magnitude of the yield impact is small when compared to differences between irrigated and rainfed production. One explanation for this result is that farmers may choose to retain a smaller proportion of land in irrigated production, so that limited groundwater pumping capacity can be used to adequately buffer crops against drought on land that remains under irrigation. Reductions in irrigated areas as an aquifer is depleted have been demonstrated theoretically [@rad2020effects; @foster2014modeling; @hrozencik2017heterogeneous; @deines2020transitions], but there has been little empirical research on farmers' irrigated area choices or how these are affected by combinations of drought exposure and groundwater scarcity.

@@

```{r ir-share, fig.cap = "The impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean in the US High Plains (Note: shaded area represents 95 percent confidence interval)", out.height = "4in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_share.pdf"))
```

@@

The overall effects of aquifer depletion on drought-related production losses is a function of both per-area yield and irrigated area changes highlighted in Sections \@ref(impact-intensive) and \@ref(impact-share). Combining both responses, Figure \@ref(fig:tot-impact) shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where weights are production area shares) for corn and soybean.

For corn, aquifer thickness does not have a significant impact on productivity when precipitation deficits are small (less than around 200mm). However, beyond this threshold, higher levels of water deficit result in noticeable and significant non-linear reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (i.e. greater than around 600mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the highest (3rd quantile) and lowest (1st quantile) aquifer thickness.

```{r tot-impact, fig.cap = "Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness", out.height = "5in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_total_impact.pdf"))
```

```{r dif-tot-impact, fig.cap = "Difference in average yield relative to the 1st aquifer thickness quantile for corn and soybean", cache = FALSE}
knitr::include_graphics(here::here("Shared/Results/g_avg_yield_dif.pdf"))
```

@@

Aquifer depletion may impact resilience of irrigated crop production to drought through two main channels: (1) a reduction in per-area yields due to the economic or physical inability to meet crop water requirements; and (2) a reduction in the area of land irrigated, resulting in reduced production due to lower yields obtained through rainfed agriculture.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(WB\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0-33.3\%)\), \([33.3-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) represents the category specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient. To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\phi_t\)) are also included and standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water balance is captured by $\sum_{k}^K \alpha_{k}\phi_{k}(WB_{i})$, where $\phi_{k}(\cdot)$ is the $k$th spline basis function and $\alpha_{k}$ is its coefficient. The impact of aquifer thickness is captured in a similar way with $\tau_{l}(\cdot)$ and $\beta_{l}$ being the $l$th spline basis function and its coefficient, respectively. State fixed effects, year fixed effects and error term are represented by $\theta_s$, $\mu_t$, and $v_{i,t}$, respectively. Unlike our regression model of the impact of water balance on yield, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects effectively eliminate the cross-sectional variation in aquifer thickness.

county FE

In the final step of our analysis, we then estimate the average crop yield for each county as a function of aquifer thickness and seasonal water deficit. The average crop yield is estimated as the average of rainfed and irrigated yields estimated on the basis of Equation \@ref(eq:eqn1), weighted by the estimated share of total acres that are irrigated in the county estimated using Equation \@ref(eq:eqn2). Comparing the estimated average crop yield across different levels of aquifer thickness for a given seasonal water deficit provides an indication of the effect of aquifer conditions on drought impacts, considering impacts of drought through changes in both per-area yields and the area of land irrigated.

@@

# More on the impact of water deficit on irrigated and average yield

@@

```{r irrigated-yield-ind-corn, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn in US High Plains with 95 percent confidence interval", out.width = "6in", cache = FALSE}
knitr::include_graphics(here::here("Shared/Results/g_yield_with_conf_corn.pdf"))
```

```{r irrigated-yield-ind-soy, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of soybean in US High Plains with 95 percent confidence interval", out.width = "6in", cache = FALSE}
knitr::include_graphics(here::here("Shared/Results/g_yield_with_conf_soy.pdf"))
```

```{r avg-yield-ind, fig.cap = "Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness (The shaded areas represent the 95 percent confidence interval)", out.width = "6.5in", cache = FALSE}
knitr::include_graphics(here::here("Shared/Results/g_avg_yield.rds"))
```

@@

# County Fixed Effects in Irrigation Share Regression

Figure \@ref(fig:variation-left) presents the histogram of aquifer thickness after demeaning them by its overall average (a), state average (b), and county average (c) to gain insights into how much variation for identification is left under no county or state fixed effects (a), state fixed effects (b), and county fixed effects (c). As you can see in the figure, while including state fixed effects do not reduce the total variation in aquifer thickness, including county fixed effects dramatically reduces its total variation. Consequently, we found that including county fixed effects results in extremely inaccurate estimates (high standard error). For this reason, we just included state fixed effects.

```{r variation-left, fig.cap = "Comparison of variations in aquifer thikcness after demeaning by overall average, county average, and state average.", out.height = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_variation.pdf"))
```

@@

```{r state-fe-with-without, fig.cap = "Comparison of estimated impact of aquifer thickness on the share of irrigated acres with and without state fixed effects", out.height = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_share_comp.pdf"))
```

# Spline basis functions and GAM estimation

@@

```{r sp-basis, fig.cap = "Example of a series of thin-plate spline basis functions", out.height = "4in", out.width = "6in"}
library(mgcv)

set.seed(83944)

# === generate data ===#
N <- 400 # number of observations
x <- seq(1, 250, length = N)
y_det <- 240 * (1 - 0.4 * exp(-0.03 * x))
e <- 25 * rnorm(N) # error
data <- data.table(x = x, y = y_det + e, y_det = y_det)

basis_data <-
  gam(y_det ~ s(x, k = 5, bs = "cr"), data = data) %>%
  predict(., type = "lpmatrix") %>%
  data.table() %>%
  .[, x := data[, x]] %>%
  melt(id.var = "x") %>%
  .[, variable := case_when(
    variable == "(Intercept)" ~ "Intercept",
    variable == "s(x).1" ~ "1st basis",
    variable == "s(x).2" ~ "2nd basis",
    variable == "s(x).3" ~ "3rd basis",
    variable == "s(x).4" ~ "4th basis"
  )] %>%
  .[, variable := factor(
    variable,
    levels= c("Intercept", "1st basis", "2nd basis", "3rd basis","4th basis")
  )]

ggplot(data = basis_data) +
  geom_line(aes(y = value, x = x)) +
  facet_grid(variable ~ .) +
  theme_bw() +
  ylab("")

```

Different coefficients (or namely weights) given to the spline basis functions result in different shapes of the imapact of $x$ on $y$. Figure \@ref(fig:ex-fx) presents two examples of two different sets of coefficients and their resulting represention of $f(x)$. In the first case, $\{\beta_1, \dots, \beta_5\} = \{60, 60, 60, 60, 60\}$. In the second case, $\{\beta_1, \dots, \beta_5\} = \{50, 20, 40, 60, 80\}$. Among all the possibilities of the combinations of the coefficients, GAM finds the set of coefficient values that fits the data best. Notice that unlike non-parametric regression, GAM is actually a linear in parameter model and estimate $\beta_k$ to fit the data best.

```{r ex-fx, fig.cap = "Example of a function represented by different coefficient values", out.height = "4in", out.width = "6in"}
library(patchwork)

g_il_1 <-
  data.table(
    variable = unique(basis_data$variable),
    coef = rep(60, 5)
  ) %>%
  .[basis_data, on = "variable"] %>%
  .[, sum(coef * value), by = x] %>%
  ggplot(data = .) +
  geom_line(aes(y = V1, x = x)) +
  theme_bw() +
  ylab("y") +
  ggtitle("Case 1")

g_il_2 <-
  data.table(
    variable = unique(basis_data$variable),
    coef = c(50, 20, 40, 60, 80)
  ) %>%
  .[basis_data, on = "variable"] %>%
  .[, sum(coef * value), by = x] %>%
  ggplot(data = .) +
  geom_line(aes(y = V1, x = x)) +
  theme_bw() +
  ylab("y")  +
  ggtitle("Case 2")

g_il_1 / g_il_2
```

>>>>

---
title: Aquifer depletion exacerbates agricultural drought losses in the US High Plains
author:
  - Taro Mieno^[Department of Agricultural Economics, University of Nebraska Lincoln, tmieno2@unl.edu]
  - Timothy Foster^[Department of WWW, University of Manchester, timothy.foster@manchester.ac.uk]
  - Shunkei Kakimoto^[Department of Applied Economics, University of Minnesota, kakim002@umn.edu]
  - Nicholas Brozovic^[Department of Agricultural Economics, University of Nebraska Lincoln, nbrozovic@nebraska.edu]
abstract: |
  Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in drought years. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
acknowledgments: |
  Funding for the research in this manuscript was provided by the United States Department of Agriculture under contract numbers OCE 58-0111-20-007 and OCE 58-0111-21-007.
bibliography: DRA.bib
csl: iopart.csl
output:
  bookdown::pdf_document2:
    extra_dependencies: ["float"]
    citation_package: natbib
    keep_tex: true
    toc: false
    includes:
      in_header: "preamble.tex"
---

@@

```{r}
#| echo: false
#| cache: false
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  fig.pos = "H",
  out.extra = ""
)

@@

@@

```{r}
test_results <- readRDS(here::here("Shared/Results/test_results_intensive.rds"))
```

Figure \@ref(fig:yield-response) shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B))^[Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \@ref(reg-conf) for estimated yield response curves with 95 percent confidence interval.]. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases. This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors [@zhu2022untangling; @li2020quantifying].

We now focus on comparing yield response to water deficitn among the three auifer thickness categories (all irrigated). First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \@ref(fig:yield-response)). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers' ability to fully meet crop water requirements during periods of more severe or extended precipitation deficits [@rad2020effects, @foster2014modeling; @hrozencik2017heterogeneous]. However, yield gradually increases as water deficit increases for counties with the highest level of aquifer thickness.

Figure \@ref(fig:yield-dif) shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quatile with the 95 percent confidence interval for corn an soybean. For corn, in years with water deficit higher than about 800 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extrement water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference is `r round(test_results[balance == 1000 & crop == "Corn" & comp_q == 3, dif_hat], digits = 2)` tonne/ha. At low water deficits (below 200mm), counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \@ref(fig:yield-response)) experience statistically lower yields than counties with the lowest level of aquifer thickness (1st quantile). This counter-intuitive result may be due to the increased possibility of over-irrigation in regions with high aquifer thickness and well capacities [@foster2019assessing; @gibson2017case; @gibson2019benchmarking]. Unlike corn, soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likey because soybean is not as water demanding as corn and even the counties the lowest aquifer thickness level can catch up with water demand in severe drought years just as well as counties with higher aquifer thickness levels.

<!-- However, it is important to highlight that impacts of aquifer thickness on yields for negative deficits is much less pronounced than that for extreme droughts (i.e. high deficits), in particular for corn. -->
```{r yield-response, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area irrigated yields of corn and soybean in US High Plains", out.width = "6.5in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_yield_intensive.pdf"))
```

```{r yield-dif, out.width = "6in", fig.cap = "The difference in yield for the counties with the second and third (highest) aquifer thickness groups relative to the first (lowest) aquifer thickness group."}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_ir_yield_dif.pdf"))
```

@@

While aquifer thickness has a statistically significant impact on per-area irrigated crop yields in drought conditions (more than 800mm water deficit), the magnitude of the yield impact is small when compared to differences between irrigated and rainfed production. One explanation for this result is that farmers may choose to retain a smaller proportion of land in irrigated production, so that limited groundwater pumping capacity can be used to adequately buffer crops against drought on land that remains under irrigation. Reductions in irrigated areas as an aquifer is depleted have been demonstrated theoretically [@rad2020effects; @foster2014modeling; @hrozencik2017heterogeneous; @deines2020transitions], but there has been little empirical research on farmers' irrigated area choices or how these are affected by combinations of drought exposure and groundwater scarcity.

@@

```{r ir-share, fig.cap = "The impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean in the US High Plains (Note: shaded area represents 95 percent confidence interval)", out.width = "5in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_share.pdf"))
```

@@

```{r}
results <- readRDS(here::here("Shared/Results/reg_results_data.rds"))

res_at_950 <-
  rbindlist(results$avg_yield_data) %>%
  .[(abs(balance - 950)) < 1, ]
```
The overall effects of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \@ref(impact-intensive) and \@ref(impact-share). Combining both responses, Figure \@ref(fig:tot-impact) shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where weights are production area shares) for corn and soybean at thee values of aquifer thickness: 12.2, 59.1, and 106.1 meters^[Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \@ref(reg-conf) for estimated average yield response curves with 95 percent confidence interval.]. While you can see similar trends as presented in Figure \@ref(fig:yield-response), the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronouced mainly dues to lower irrigated area for lower aquifer thickness level (rainfed production has a larger share).

Figure \@ref(fig:dif-tot-impact) shows the difference in average per-area production of the 59.1m and 106.1m aquifer thickness leves relative to 12.2m with the 95 percent confidence interval for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when precipitation deficits are small (less than around 200mm and 300mm for 59.1m and 106.1m aquifer thickness, respectively). However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (i.e. greater than around 800mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the highest (3rd quantile) and lowest (1st quantile) aquifer thickness. Further, notice that the decline in average per-area production at a sesonal water deicit of 950mm is not linear. Going from 106.1m aquifer thickness to 59.1m (decline of 47m) would result in `r round(res_at_950[sat == 348, avg_yield] - res_at_950[sat == 194, avg_yield], digits = 2)` tonne/ha of production loss for corn. However, going from 59.1m aquifer thickness to 12.2m (decline of 46.9m) would result in `r round(res_at_950[sat == 194, avg_yield] - res_at_950[sat == 40, avg_yield], digits = 2)` tonne/ha of production loss for corn. That is the same degree of decline is much more damaging when the decline is from a low aquifer thickness level thant a higher aquifer thickness level. This is primarily due to the non-linear impact of aquifer thickness on the share of irrigated corn acres as shown in Figure \@ref(fig:ir-share). When aquifer thickness level is high enough (e.g., 100m), a small decline does not reduce the share of irrigated acres. However, when aquifer thickness level is low (e.g., 40m), a small decline does reduce the share of irrigated acres.

```{r tot-impact, fig.cap = "Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness", out.width = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_total_impact.pdf"))
```

```{r dif-tot-impact, fig.cap = "Difference in average yield relative to the 1st aquifer thickness quantile for corn and soybean", out.width = "6in", cache = FALSE}
knitr::include_graphics(here::here("Shared/Results/g_avg_yield_dif.pdf"))
```

@@

Aquifer depletion may impact resilience of irrigated crop production to drought through two main channels: (1) a reduction in per-area irrigated yields due to the economic or physical inability to meet crop water requirements; and (2) a reduction in the area of land irrigated, resulting in reduced production due to lower yields obtained through rainfed agriculture.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(WB\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0-33.3\%)\), \([33.3-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) represents the category specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \@ref(spline-basis) for further explanations on how spline-based regression works). To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\phi_t\)) are also included and standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water balance is captured by $\sum_{k}^K \alpha_{k}\phi_{k}(WB_{i})$, where $\phi_{k}(\cdot)$ is the $k$th spline basis function and $\alpha_{k}$ is its coefficient. The impact of aquifer thickness is captured in a similar way with $\tau_{l}(\cdot)$ and $\beta_{l}$ being the $l$th spline basis function and its coefficient, respectively. State fixed effects, year fixed effects and error term are represented by $\theta_s$, $\mu_t$, and $v_{i,t}$, respectively. One of the limitations here is that we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable unlike our regression model of the impact of water balance on yield. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects effectively eliminate the cross-sectional variation in aquifer thickness. The trade-off here is tighter controls on potential unobservables that may be correlated with aquifer thickness (see Appendix \@ref(county-fe) for more discussions).

In the final step of our analysis, we then estimate the average crop yield as a function of aquifer thickness and seasonal water deficit. The average crop yield is estimated as the average of rainfed and irrigated yields estimated on the basis of Equation \@ref(eq:eqn1), weighted by the estimated share of total acres that are irrigated in the county estimated using Equation \@ref(eq:eqn2). Comparing the estimated average crop yield across different levels of aquifer thickness for a given seasonal water deficit provides an indication of the effect of aquifer conditions on drought impacts, considering impacts of drought through changes in both per-area irrigated yields and the area of land irrigated. Its standard error is estimated using bootstrap clustered by county.

@@

# More on the impact of water deficit on irrigated and average yield {#reg-conf}

@@

```{r irrigated-yield-ind-corn, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area irrigated yields of corn in US High Plains with 95 percent confidence interval", out.width = "6in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_yield_with_conf_corn.pdf"))
```

```{r irrigated-yield-ind-soy, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area irrigated yields of soybean in US High Plains with 95 percent confidence interval", out.width = "6in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_yield_with_conf_soy.pdf"))
```

```{r avg-yield-ind, fig.cap = "Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness (The shaded areas represent the 95 percent confidence interval)", out.width = "6.5in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_avg_yield.pdf"))
```

@@

# County Fixed Effects in Irrigation Share Regression {#county-fe}

Figure \@ref(fig:variation-left) presents the histogram of aquifer thickness after demeaning them by its overall average (A), state average (B), and county average (C) to gain insights into how much variation for identification is left under no county or state fixed effects (A), state fixed effects (B), and county fixed effects (C). As you can see in the figure, while including state fixed effects do not reduce the total variation in aquifer thickness, including county fixed effects dramatically reduces its total variation. Consequently, we found that including county fixed effects results in extremely inaccurate estimates (high standard error). For this reason, we just included state fixed effects.

```{r variation-left, fig.cap = "Comparison of variations in aquifer thikcness after demeaning by overall average, county average, and state average.", out.width = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_variation.pdf"))
```

@@

```{r state-fe-with-without, fig.cap = "Comparison of estimated impact of aquifer thickness on the share of irrigated acres with and without state fixed effects", out.width = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_share_comp.pdf"))
```

# Spline basis functions and GAM estimation {#spline-basis}

```{=Tex}
\setcounter{figure}{0}
\renewcommand{\thefigure}{C.\arabic{figure}}
```

@@

```{r sp-basis, fig.cap = "Example of a series of thin-plate spline basis functions", out.width = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_basis.pdf"))
```

Different coefficients (or namely weights) given to the spline basis functions result in different shapes of the imapact of $x$ on $y$. Figure \@ref(fig:ex-fx) presents two examples of two different sets of coefficients and their resulting represention of $f(x)$. In the first case, $\{\beta_1, \dots, \beta_5\} = \{60, 60, 60, 60, 60\}$. In the second case, $\{\beta_1, \dots, \beta_5\} = \{20, -20, -40, 40, 60\}$. Among all the possibilities of the combinations of the coefficients, GAM finds the set of coefficient values that fits the data best. Notice that unlike non-parametric regression, GAM is actually a linear in parameter model and estimate $\beta_k$ to fit the data best.

```{r ex-fx, fig.cap = "Example of a function represented by different coefficient values", out.width = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_il_y.pdf"))
```
